---
echo: true
---

# Showcase: Importing Sentinel data

## Download data

To download Sentinel data:

1. Go to the [Copernicus Browser](https://browser.dataspace.copernicus.eu/) website
2. Login using your credentials
3. Go to your area of interest
4. Click on the "Search" Tab in the left panel
5. In Data Source, select *Sentinel 2* → *L2A*
6. Choose your desired time frame
7. Click on the "Search" button
8. From the search results, download one or two scenes


## Copernicus SAFE Format I


- Sentinel 2 data is provided in SAFE format. This is a zipped file that contains the data in separate JP2 files
- The Sentinel-SAFE format wraps a folder containing image data in a binary data format and product metadata in XML. 
- This includes:
  - A 'manifest.safe' file which holds the general product information in XML.
  - Subfolders for measurement datasets containing image data in various binary formats.
  - A preview folder containing 'quicklooks' in PNG format, Google Earth overlays in KML format and HTML preview files.
  - An annotation folder containing the product metadata in XML as well as calibration data.
  - A support folder containing the XML schemes describing the product XML.



## Copernicus SAFE Format II

The SAFE folder above includes the following content:

```{r}
#| echo: false

library(data.tree)
library(tidyr)
library(dplyr)

tibble(full = list.files("data-large/S2B_MSIL2A_20240915T102559_N0511_R108_T32TMS_20240915T131207.SAFE/", recursive = TRUE, full.names = TRUE)) |> 
  mutate(
    basen = basename(full),
    dirn = dirname(full)
  ) |> 
  separate_wider_delim(basen,delim = "_", names = paste0("d",1:7),too_few = "align_start",cols_remove = FALSE) |> 
  group_by(dirn, d1,d2) |> 
  mutate(
    n = row_number(),
    basen = ifelse(n == 3, "... (etc)", basen)
    ) |> 
  filter(n <= 3) |> 
  ungroup() |> 
  transmute(paths = paste(dirn, basen, sep = "/")) -> fi

as.Node(fi,pathName = "paths")

```


The data we are interested in is in the `IMG_DATA` folder. This folder contains the data in three different resolutions, where each band is a separate file *jp2* File. 

```{r}
library(terra)
library(dplyr)

s2_files <- list.files("data-large/S2B_MSIL2A_20240915T102559_N0511_R108_T32TMS_20240915T131207.SAFE/GRANULE/L2A_T32TMS_A039316_20240915T102803/IMG_DATA/R60m/", "\\.jp2$", full.names = TRUE)
```


- We can import all `jp2` files into a `rast` object in a singe `rast()` command. 
- Before we do this, let's filter the data for the bands we are interested in (`B01`...`B12`)
- To add reasonable `names` to the SpatRaster object, we can use `str_split_fixed` to extract the relevant information from the file names.

```{r}
library(stringr)

# selecting only the bands we are interested in:
s2_files <- s2_files[str_detect(s2_files, "B\\d{2}")]

s2 <- rast(s2_files)

# Extracting the band names from the file names and adding these
names(s2) <- str_split_fixed(names(s2), "_",4)[,3]

names(s2)
```

- Test if the data is loaded correctly:

```{r}
plot(s2[[1]])
```

- Note how the values range from 0 to > 15000. It seems that the values are not scaled yet (see @sec-data-types)
- If `scale` and `offset` values were set using `GDAL` flags, `terra` would automatically apply these values
- However, it seems that we have to apply these values manually: [this website](https://sentinel.esa.int/en/web/sentinel/technical-guides/sentinel-2-msi/level-1c/product-formatting), writes the following:

> The transformation of reflectances in 16 bit integers is performed according to the following equation:
> $$\text{L1C\_DN} = \rho \times \text{QUANTIFICATION\_VALUE} - \text{RADIO\_ADD\_OFFSET}$$
> The L1C product's metadata includes the values for the QUANTIFICATION_VALUE and RADIO_ADD_OFFSET.

- This information is stored in the file `MTD_MSIL2A.xml`:
- This file contains information about the bands, the scale factor, and the offset values. 
- If we search for `QUANTIFICATION_VALUE`, we find the following information:

```{r}
#| echo: false

meta <- readLines("data-large/S2B_MSIL2A_20240915T102559_N0511_R108_T32TMS_20240915T131207.SAFE/MTD_MSIL2A.xml") 

cat(meta[str_detect(meta, "QUANTIFICATION_VALUE")], sep = "\n")
```

- If we search for `ADD_OFFSET`, we find the following information:


```{r}
#| echo: false

cat(meta[str_detect(meta, "ADD_OFFSET")], sep = "\n")
```

<!-- this does not work, since I would need offset to be added first, and then scale (?) -->
<!-- - Rather than manually converting our data to reflectance values, we can use the function `scoff` -->
<!-- - This function takes a *two-column matrix with scale (first column) and offset (second column) for each layer* -->
<!-- - `scale` and `offset` are applied to the original values as follows: -->

<!--   > `value <- value * scale + offset` -->

- We can now use the `scale` (10'000) and `offset` (-1'000) values to convert the data to reflectance values. 
- We can then use `hist` to get a histogram of the values in each band
- Note that most bands have values > 1. 
  - We have to decide how to handle these values
  - One option is to simply clip the values to the desired range

```{r}
s2b <- (s2 - 1000)/10000

hist(s2b)
```


```{r}
# function to clip values to the desired range
force_minmax <- \(x, min = 0, max = 1){
  x[x<min] <- min
  x[x>max] <- max
  x
}
```

To apply this function to all bands, we can use the `app` function. 

```{r}
s2c <- app(s2b, force_minmax)

# If we now check the min/max values, we see that all values are between 0 and 1
minmax(s2c)
```

To test if we processed the data correctly, let's create a True Color image using the bands `B04`, `B03`, and `B02`. 

```{r}
plotRGB(s2c,r = 4, g = 3, b = 2, stretch = "histogram", smooth = FALSE)
```

```{r}
# To export the R,G,B bands to a True Color Geotiff: 
writeRaster(s2c[[c(4,3,2)]], "data-out/sentinel-rgb.tif", overwrite = TRUE)
```


To calculate NDVI, we can create our custom function and apply it to the `s2c` object using `app`:

```{r}
ndvi <- \(x){(x[5]-x[4])/(x[5]+x[4])}

s2_ndvi <- app(s2c, ndvi)

library(tmap)
tm_shape(s2_ndvi) + 
  tm_raster(style = "cont",midpoint = 0) + 
  tm_layout(legend.outside = TRUE)
```


