---
execute:
  echo: true                 # ← set to true to show sample solution
  collapse: true
code-fold: true
code-summary: "Sample Solution"
---

## Tasks

:::{.callout-important}
- **Try this without AI first** - struggle builds understanding. 
- Ask if you are struggling. Asking builds understanding as well!
:::

## Task 1

The following command returns the path to a tif file on your hard drive:

```{r}
#| eval: false
#| echo: true
#| code-fold: false
system.file("ex/elev.tif", package="terra")
```
   
Use this path to import the tif file using `rast()`, store it as `r`.


```{r}
library(terra)
r <- system.file("ex/elev.tif", package="terra") |> 
  rast()
```


   
## Task 2

Explore this object:

   - Determine the minimum and maximum elevation values
   - Make a static map using base plot and `tmap`
   - Make an interactive map using tmap (`tmap_mode("view")`)
   - Using tmap, explore different [styles](https://r-tmap.github.io/tmap/reference/tm_raster.html#arg-style) and [palettes](https://r-tmap.github.io/tmap/reference/tm_raster.html#arg-palette). 


```{r}
# Simply looking at the metadata will show min/max values
r

library(tmap)

tm_shape(r) +
  tm_raster()

```

## Task 3

The following command returns the path to a tif file on your hard drive:

```{r}
#| eval: false
#| echo: true
#| code-fold: false
system.file("tif/L7_ETMs.tif",package = "stars")
```

Use this path to import the tif file using `rast()`, store it as `l7`.

```{r}
l7 <- system.file("tif/L7_ETMs.tif",package = "stars") |> 
  rast()
```


## Task 4

Explore this object:

- Spot the differences to the object `r`
- Plot the available layers individually
- Rename the layers to: `c("B", "G", "R", "NIR", "SWIR", "MIR")` (see [here](https://r-spatial.github.io/stars/reference/L7_ETMs.html))


## Task 5

Select the Red Green and Blue bands to create a true color map (`plotRGB` and `tm_rgb`)

```{r}
names(l7) <- c("B", "G", "R", "NIR", "SWIR", "MIR")
```

## Task 6

Select the NIR, Green and Blue bands to create a false color composite

:::{.notes}
- Allow 15-20 minutes for these exercises
- Help students with band selection syntax - this is often confusing initially
- Encourage experimentation with different band combinations
- Point out how vegetation appears differently in true vs false color
- This hands-on experience reinforces the theoretical concepts
:::

```{r}
#| eval: false


tm_shape(l7[[c("NIR","G","B")]]) + tm_rgb()

```



 

