---
execute:
  echo: true                 # ← set to true to show sample solution
  collapse: true
code-fold: true
code-summary: "Sample Solution"
---

# Tasks

:::{.callout-important}
- **Try this without AI first** - struggle builds understanding. 
- Ask if you are struggling. Asking builds understanding as well!
:::

## Task 1

Import the `Produc` dataset (from the package `plm`) using the following code:

```{r}
#| echo: true
#| code-fold: false
data(Produc, package = "plm")
```

## Task 2

 Convert the data from long to wide

```{r}
#| output: false
library(plm)
library(dplyr)
library(tidyr)
library(tibble)
data("Produc", package = "plm")

Produc_wide <- Produc |> 
  select(year, state, unemp) |> 
  pivot_wider(names_from = state, values_from = unemp) |> 
  column_to_rownames("year")


```



## Task 3

Convert the data to a matrix

```{r}
Produc_matrix <- as.matrix(Produc_wide)
```


## Task 4

Calculate mean unemployment first using the wide dataframe, then the new matrix. Compare the speed difference.

```{r}
library(bench)
bm <- bench::mark(
    matrix = mean(Produc_matrix, na.rm = TRUE),
    dataframe = mean(Produc_wide |> unlist(),na.rm = TRUE),
    filter_gc = FALSE
)

bm$total_time[2]/bm$total_time[1]

```





